---
title: "sprtt"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  # fig.path = "man/figures/README-",
  out.width = "100%"
)
options(tibble.print_min = 5, tibble.print_max = 5)
```

<!-- badges: start -->

[![codecov](https://codecov.io/gh/MeikeSteinhilber/sprtt/branch/main/graph/badge.svg?token=IQHTDTRBAW)](https://codecov.io/gh/MeikeSteinhilber/sprtt) [![R-CMD-check-small](https://github.com/MeikeSteinhilber/sprtt/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/MeikeSteinhilber/sprtt/actions/workflows/R-CMD-check.yaml) [![R-CMD-check-big](https://github.com/MeikeSteinhilber/sprt/workflows/R-CMD-check/badge.svg)](https://github.com/MeikeSteinhilber/sprtt/actions/workflows/check-full.yaml)

<!-- badges: end -->

## Overview

**sprtt** provides Sequential Probability Ratio tests using the associated t-statistic:

-   `seq_ttest()` calculates the sequential test statistic.
-   `print_seq_ttest()` visualizes the test statistic.

## Installation

```{r, eval = FALSE}
# Code for installation

```

### Development version

To get a bug fix or to use a feature from the development version, you can install the development version of **sprtt** from GitHub.

    # install.packages("devtools")
    devtools::install_github("sprtt")

## Usage

```{r, eval = FALSE}
# Test Code

```
