---
title: "sprtt"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  # fig.path = "man/figures/README-",
  out.width = "100%"
)
options(tibble.print_min = 5, tibble.print_max = 5)
```

<!-- badges: start -->

[![codecov](https://codecov.io/gh/MeikeSteinhilber/sprtt/branch/main/graph/badge.svg?token=IQHTDTRBAW)](https://codecov.io/gh/MeikeSteinhilber/sprtt)

[![R-CMD-check-windows-macOs](https://github.com/MeikeSteinhilber/sprtt/workflows/R-CMD-check-windows-macOs/badge.svg)](https://github.com/MeikeSteinhilber/sprtt/actions)

[![R-CMD-check-ubuntu20.04](https://github.com/MeikeSteinhilber/sprtt/workflows/R-CMD-check-ubuntu20.04/badge.svg)](https://github.com/MeikeSteinhilber/sprtt/actions)

[![R-CMD-check-ubuntu16.04](https://github.com/MeikeSteinhilber/sprtt/workflows/R-CMD-check-ubuntu16.04/badge.svg)](https://github.com/MeikeSteinhilber/sprtt/actions)

<!-- badges: end -->

## Overview

**sprtt** provides Sequential Probability Ratio Tests (SPRT) using the associated t-statistic:

-   `seq_ttest()` calculates the sequential test statistic.
-   `print_seq_ttest()` visualizes the test statistic.

## Installation

```{r, eval = FALSE}
# Code for installation

```

### Development version

To get a bug fix or to use a feature from the development version, you can install the development version of **sprtt** from GitHub.
```{r, eval=FALSE}
# install.packages("devtools")
devtools::install_github("MeikeSteinhilber/sprtt")
```


   

## Usage

```{r, eval = FALSE}
# Test Code

```
