---
title: "Sequential One-Way ANOVA"
author: "Meike Snijder-Steinhilber"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 4
description: >
  This vignette describes the sequential one-way ANOVA.
vignette: >
  %\VignetteIndexEntry{Sequential One-Way ANOVA}
  %\VignetteEncoding{UTF-8}{inputenc}
  %\VignetteEngine{knitr::rmarkdown}
bibliography: references.bib
csl: "apa.csl"
---

```{r, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE
)
```

## What is the sequential one-way ANOVA?

The sequential one-way fixed effects ANOVA is a sequential hypothesis test based on the Sequential Probability Ratio Test (SPRT) framework [@wald1947].
It extends SPRTs to the comparison of two or more independent groups and can be used as an efficient alternative to the classical fixed-sample one-way ANOVA.
For detailed information, see @steinhilber2024.
For a general introduction to SPRTs, see the vignette `vignette("sprts")`.

**Note:** The repeated measures ANOVA is not yet implemented in the `sprtt` package.



## Sequential One-Way ANOVA

Analysis of variance (ANOVA) is widely used to compare means across multiple groups.
Traditional fixed-sample ANOVAs require an a priori power analysis to determine the necessary sample size, which often results in large samples, especially when expected effect sizes are small [@steinhilber2024].
Given the prevalence of small to medium effects in psychology [@funder2019; @richard2003], many studies end up underpowered because the required sample sizes exceed available resources [@button2013; @szucs2017].

The sequential one-way ANOVA provides an efficient alternative by applying the SPRT framework to comparisons of $k$ groups.
Instead of collecting a fixed number of observations, data are collected sequentially and the test evaluates evidence after each step until a decision is reached [@steinhilber2024].

### Hypotheses

The sequential one-way ANOVA tests the following hypotheses, specified in terms of Cohen's $f$:


```{=tex}
\begin{aligned}
H_0 &: f = 0 \\
H_1 &: f = f_{\text{exp}}, \quad (f_{\text{exp}} > 0)
\end{aligned}
```



Cohen's $f$ is defined as:

$$f = \frac{\sigma_m}{\sigma}$$

where $\sigma$ is the common within-population standard deviation and $\hat{\sigma}_m = \sqrt{\frac{\sum_{i=1}^{k}(m_i - \bar{m})^2}{k}}$, with $m_i$ being the mean of group $i$ and $\bar{m}$ the overall mean across all $k$ groups. In other words, $f$ captures the spread of group means relative to the common within-group variability.

### The $F$ Statistic

At the $n$-th sequential step, the $F$ statistic is calculated from $k$ groups with $n$ observations each (total $N = k \cdot n$, with $n, k \geq 2$):

$$F_n = \frac{SS_{\text{effect},n} / df_1}{SS_{\text{residual},n} / df_{2,n}}$$

with

$$SS_{\text{effect},n} = \sum_{i=1}^{k} n(\bar{x}_i - \bar{x})^2$$

$$SS_{\text{residual},n} = \sum_{i=1}^{k} \sum_{j=1}^{n} (x_{i,j} - \bar{x}_i)^2$$

$$df_1 = k - 1 \qquad \text{and} \qquad df_{2,n} = N - k$$

where $\bar{x}$ is the overall mean, $\bar{x}_i$ is the mean of group $i$, and $x_{i,j}$ is the $j$-th observation in group $i$ [@wetherill1986; @steinhilber2024].

### The Likelihood Ratio

The likelihood ratio at step $n$ is defined as the ratio of the likelihood under $H_1$ to the likelihood under $H_0$.
Using Cox's theorem [@cox1952], it is sufficient to compute this ratio only for the current $F_n$ statistic rather than the entire sequence of observations:

$$\text{LR}_n = \frac{f(F_n \mid df_1,\; df_{2,n},\; \Delta_{1n})}{f(F_n \mid df_1,\; df_{2,n})}$$

The numerator is the density of a **non-central** $F$ distribution with non-centrality parameter $\Delta_{1n}$, and the denominator is the density of a **central** $F$ distribution (i.e., $\Delta = 0$ under $H_0$). The non-centrality parameter is linked to Cohen's $f$ via:

$$\Delta_1 = f_{\text{exp}}^2 \cdot N$$

### Decision Rule

The sequential ANOVA applies the standard SPRT decision boundaries:

- If $\text{LR}_n \geq A = \frac{1-\beta}{\alpha}$: Stop and accept $H_1$
- If $\text{LR}_n \leq B = \frac{\beta}{1-\alpha}$: Stop and accept $H_0$
- If $B < \text{LR}_n < A$: Continue collecting data

### Efficiency and Robustness

Simulations with $k = 4$ groups demonstrate that the sequential ANOVA is substantially more efficient than fixed-sample designs: in 87% of cases the sequential sample was smaller than the fixed sample, with an average sample size reduction of 58%.
Efficiency gains are particularly pronounced for small expected effect sizes [@steinhilber2024].

Robustness analyses showed that the sequential and fixed ANOVA behave similarly under assumption violations.
Both procedures are robust to non-normal data (simulated using Gaussian mixture distributions) and to mildly unequal variances or group sizes.
The most critical scenario is a combination of unbalanced group sizes and unequal variances, specifically when smaller groups have larger variances—in this case, $\alpha$ error rates can be inflated in both the sequential and the fixed ANOVA.

As with all sequential procedures, effect size estimates from individual sequential ANOVAs are conditionally biased (see Section "The Bias-Efficiency Tradeoff").
However, the weighted average across studies closely approximates the true population effect size [@steinhilber2024].

The `seq_anova()` function in the `sprtt` package implements the sequential one-way fixed effects ANOVA described here.

## How to use `seq_anova()`

In the first step, we simulate data that we can analyze.
In a real world example we use the data that are coming in from the data collection.

```{r example-0.1}
set.seed(333)
# generate data with a medium effect -------------------------------------------
data <- sprtt::draw_sample_normal(
                k = 3,
                f = 0.25,
                max_n = 22)
```

We can calculate the sequential ANOVA for the first time, after we have 2 data points in each group.

```{r example-0.2}
# calculate the SPRT -----------------------------------------------------------
anova_results <- sprtt::seq_anova(
                          y~x,
                          f = 0.25,
                          data = data[1:6,],
                          verbose = FALSE)
anova_results

# access the decision ----------------------------------------------------------
anova_results@decision
```

The decision is, that we have to continue the data collection.
In the best case, we calculate the SPRT after each new data point.

Lets assume, we have now reached a later stage in the data collection -- in this scenario we have collected 20 data points.

```{r example-0.3}
# calculate the SPRT -----------------------------------------------------------
anova_results <- sprtt::seq_anova(
                          y~x,
                          f = 0.25,
                          data = data[1:20,],
                          verbose = FALSE)
anova_results

# access the decision ----------------------------------------------------------
anova_results@decision
```

We still got the decision to continue the data collection, so we do that.

```{r}
# calculate the SPRT -----------------------------------------------------------
anova_results <- sprtt::seq_anova(
                          y~x,
                          f = 0.25,
                          data = data,
                          verbose = TRUE)
anova_results

# access the decision ----------------------------------------------------------
anova_results@decision

# acess the LR -----------------------------------------------------------------
anova_results@likelihood_ratio
```

With a sample size of $N = $ `r anova_results@total_sample_size`, we now have reached the decision to `anova_results@decision`.
Thus, we stop the data collection.

## How to plot the ANOVA results

In order to plot the likelihood progression, we have to calculate all the sequential steps that have happened before we reached the current sample size.
As this is only necessary for plotting and increases the run time of the function, these calculations are only done, if the function argument `plot=TRUE` is set.


### Scenario 1: Perfect data

In this case, we have data that are perfectly balanced and in a perfect sampling order.
Here, we can use the default value of the 'plot' argument 'single' or we can choose 'balanced'.
See the function documentation of `seq_anova()` for a description of each function argument

```{r example-1}
set.seed(333)
data <- sprtt::draw_sample_normal(3, f = 0.25, max_n = 22)

# calculate the SPRT -----------------------------------------------------------
anova_results <- sprtt::seq_anova(y~x, f = 0.25,
                                  data = data, plot = TRUE)
anova_results <- sprtt::seq_anova(y~x, f = 0.25,
                                  data = data, plot = TRUE,
                                  seq_steps = "single")
anova_results <- sprtt::seq_anova(y~x, f = 0.25,
                                  data = data, plot = TRUE,
                                  seq_steps = "balanced")

# plot the results -------------------------------------------------------------
sprtt::plot_anova(anova_results)
```


### Scenario 2: Unbalanced data in an imperfect order

In this case, we have a data set with unbalanced sample sizes between the groups and the data points are not in a perfect order.
Because the order is not perfect, it does not make sense to use the 'balanced' option.
Because the first data points (2*k_groups) are not equally distributed between the groups (some groups have less than 2 data points), the option 'single' would output an error.

**Thus, we need to define the sequential steps ourselves.**

Here, we start later (after 12 data points) but then calculate the LR after every single data point.

```{r example-2}
set.seed(333)
data <- sprtt::draw_sample_normal(3, f = 0.25, max_n = 37, sample_ratio = c(1,1,2))
data <- data[sample(nrow(data)),] # destroy the perfect order of the data

# calculate the SPRT -----------------------------------------------------------
anova_results <- sprtt::seq_anova(
                          y~x,
                          f = 0.25,
                          data = data,
                          plot = TRUE,
                          seq_steps = 12:nrow(data)) # we start with the first 12 data points instead of the first 6

# plot the results -------------------------------------------------------------
sprtt::plot_anova(anova_results,
                 labels = TRUE,
                 position_labels_x = 0.2,
                 position_labels_y = 0.2,
                 position_lr_x = 100,
                 position_lr_y = 1.8,
                 font_size = 20,
                 line_size = 1,
                 highlight_color = "steelblue"
                 )
```


## References
