---
title: "Sample Size Planning for Sequential ANOVAs"
author: "Meike Snijder-Steinhilber"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 4
description: >
  This vignette describes how to plan sample sizes for sequential ANOVAs using simulation-based approaches.
vignette: >
  %\VignetteIndexEntry{Sample Size Planning for Sequential ANOVAs}
  %\VignetteEncoding{UTF-8}{inputenc}
  %\VignetteEngine{knitr::rmarkdown}
bibliography: references.bib
csl: "apa.csl"
---

<style>

/* Custom alert box styling */
.alert-info {
  background-color: white !important;
  border: 2px solid #2c3e50 !important;
  border-left: 5px solid #2c3e50 !important;
  color: #2c3e50 !important;
  margin-top: 1.5em !important;
  margin-bottom: 1.5em !important;
  border-radius: 4px !important;
  padding-left: 50px !important;
  position: relative !important;
}

.alert-info::before {
  content: "🛈";
  position: absolute;
  left: 0;
  top: 0;
  bottom: 0;
  width: 40px;
  background-color: #2c3e50;
  color: white;
  display: flex;
  align-items: center;
  justify-content: center;
  font-size: 1.3em;
  border-radius: 2px 0 0 2px;
}

.alert-warning {
  background-color: white !important;
  border: 2px solid #18bc9c !important;
  border-left: 5px solid #18bc9c !important;
  color: #2c3e50 !important;
  margin-top: 1.5em !important;
  margin-bottom: 1.5em !important;
  border-radius: 4px !important;
  padding-left: 50px !important;
  position: relative !important;
}

.alert-warning::before {
  content: "⚠";
  position: absolute;
  left: 0;
  top: 0;
  bottom: 0;
  width: 40px;
  background-color: #18bc9c;
  color: white;
  display: flex;
  align-items: center;
  justify-content: center;
  font-size: 1.3em;
  border-radius: 2px 0 0 2px;
}

</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = FALSE,
  message = TRUE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

## Why Sample Size Planning Matters

Sample size planning for sequential tests like the sequential ANOVA is, in a strict statistical sense, unnecessary -- the final sample size is determined by the data and remains unknown beforehand.
Data collection continues until either the upper or lower decision boundary is reached.

However, what holds true in theory often diverges from practical constraints.
Resource planning is essential, as there is a substantial difference between collecting 100 versus 1,000 observations.
To address these practical concerns, the **sprtt** package provides the `plan_sample_size()` function, which generates HTML reports summarizing simulation results for sequential ANOVAs.

Although the exact final sample size cannot be known in advance, simulation-based recommendations can guide researchers on:

- The typical amount of data required to reach a decision
- The upper limit of resources recommended to achieve a specified decision rate

### The Challenge of Limited Resources

While the decision boundaries of the sequential ANOVA control Type I ($\alpha$) and Type II ($\beta$) errors in the long run -- and consequently maintain the desired power ($1-\beta$) -- a new consideration emerges when researchers face resource constraints.
When the maximum affordable sample size is reached before a decision boundary is crossed, this results in a **non-decision**.

Importantly, the non-decision rate depends directly on the maximum sample size a researcher can collect.
This introduces a new metric: the **decision rate** (the chance to reach a decision) given resource limitations.

### Non-Decisions vs. Null Hypothesis Acceptance

While a non-decision are undesirable, it represents a crucial conceptual distinction from accepting the null hypothesis.
SPRTs like the sequential ANOVA differentiate between stopping the data collection to accepting the null hypothesis and the case where more data/ evidence is required to stop.

<!-- ::: {.alert-info} -->
<!-- Consider a common situation in traditional fixed-design studies: a non-significant *p*-value is achieved in a setting where the sample size was planned for a medium effect size of interest with a Power of 80%. -->
<!-- In such cases, it remains unclear whether the non-significant result constitutes evidence for the null hypothesis or simply reflects a lack of evidence altogether. -->
<!-- ::: -->



## The `plan_sample_size()` Function

The `plan_sample_size()` function generates interactive HTML reports for sample size planning by querying a large simulation database.
Reports include recommended maximum sample sizes, expected sample sizes, early stopping probabilities, power curves, and comparisons to traditional ANOVA designs.

### Pre-computed Simulation Database

To make sample size planning fast and accessible, **sprtt** includes access to extensive simulation results. These simulations were conducted by:

1. Generating thousands of datasets for each combination of parameters
2. Running sequential ANOVAs on each dataset
3. Recording when each test stopped
4. Aggregating these results to get key summary statistics that guide the sample size planning

This simulation database is stored externally to keep the package installation size small.
The data are downloaded automatically on first use and cached locally for future sessions.

```{r, echo=FALSE, results='asis'}
cat(sprintf(
  'Source code of the simulation database: <a href="https://github.com/MeikeSteinhilber/sprtt_plan_sample_size" target="_blank" rel="noopener"><img src="https://img.shields.io/badge/GitHub-MeikeSteinhilber/sprtt__plan__sample__size-blue?logo=github" alt="GitHub" style="vertical-align:middle;"></a></span></p>',
  sprtt_cit
))
```


### Supported Design Parameters

The simulation database covers a range of common experimental designs:

- **Expected effect sizes (Cohen's *f*):** 0.10, 0.15, 0.20, 0.25, 0.30, 0.35, 0.40
- **Number of groups:** 2, 3, 4
- **Desired power:** 0.80, 0.90, 0.95
- **Alpha level:** 0.05


---

## Getting Started

### Installation and Setup

First, ensure the **sprtt** package is installed and loaded:
```{r installation}
# Install from CRAN (if needed)
install.packages("sprtt")

# Load the package
library(sprtt)
```

### Your First Sample Size Report


Let's walk through a practical example.
Imagine you're planning a study to compare three groups. You want to detect medium-sized effects (Cohen's *f* = 0.25) or larger, and you're working with some specific constraints.

First, you'll set your alpha level to 0.05, the standard threshold that ensures you can trust decisions to reject the null hypothesis and minimize Type I errors.
You also want high statistical power so you can trust decisions to accept the null hypothesis and minimize Type II errors.
However, given your limited resources, you're willing to accept a 20% non-decision rate.
This means that when you *do* reach a decision, which will happen 80% of the time, you can trust it, whether you're accepting or rejecting the null hypothesis $\alpha = \beta = .05$.

Now let's see how to generate a sample size planning report for this scenario:
```{r first-report}
plan_sample_size(
  f_expected = 0.25,   # Expected effect size
  k_groups = 3,        # Number of groups
  power = 0.95,        # Desired power
  decision_rate = 0.80 # desired percentage of decisions 
)
```

When you run this code for the first time, several things happen:

1. **Data download:** The simulation database (~70 MB) is downloaded from GitHub and saved to your local cache directory. This is a one-time operation.
2. **Report generation:** An HTML report is created in your temporary directory.
3. **Browser launch:** The report automatically opens in your default web browser (if running interactively).

The entire process typically takes a couple of seconds for the initial download, then just a few seconds for generating the subsequent report.



### Function Parameters

| Parameter | Type | Default | Description |
|-----------|------|---------|-------------|
| `f_expected` | numeric | *required* | Expected standardized effect size (Cohen's *f*). Must be one of: 0.10, 0.15, 0.20, 0.25, 0.30, 0.35, or 0.40. |
| `k_groups` | integer | *required* | Number of groups to compare. Must be 2, 3, 4, or 5. |
| `power` | numeric | 0.95 | Desired statistical power. Must be 0.80, 0.90, or 0.95. |
| `output_dir` | character | `tempdir()` | Directory where the HTML report will be saved. |
| `output_file` | character | `"sprtt-report-sample-size-planning.html"` | Filename for the generated report. |
| `open` | logical | `interactive()` | Whether to open the report in your browser after generation. Set to `FALSE` for batch processing. |
| `overwrite` | logical | FALSE | Whether to overwrite an existing file with the same name without prompting. |

### Return Value

The function invisibly returns the full path to the generated HTML file as a character string.
This is useful if you want to programmatically access or move the file after generation.

### Input Validation

The function validates all inputs before generating the report.
If you specify a parameter value that doesn't exist in the simulation database, you'll receive an informative error message listing the available options. For example:
```{r error-example, eval=FALSE}
# This will produce an error:
plan_sample_size(f_expected = 0.22, k_groups = 3)
#> Error: `f_expected` = 0.22 is not available. 
#> Please choose one of 0.1, 0.15, 0.2, 0.25, 0.3, 0.35, or 0.4
```



## Practical Use cases

### Case 1: Comparing Different Effect Sizes

Effect size has a large impact on required sample size.
Here's how to generate reports for different scenarios:
```{r example-effect-sizes}
# Planning for a small effect (f = 0.15)
plan_sample_size(f_expected = 0.15, k_groups = 3, power = 0.95)

# Planning for a large effect (f = 0.40)
plan_sample_size(f_expected = 0.40, k_groups = 3, power = 0.95)
```



### Case 2: Saving Reports to a Specific Location

By default, reports are saved to a temporary directory. For reports you want to keep, specify a custom location:
```{r example-custom-location}
plan_sample_size(
  f_expected = 0.25,
  k_groups = 4,
  output_dir = "~/Documents/research/sample_size_planning",
  output_file = "study1_anova_power.html",
  open = TRUE
)
```

This is particularly useful when preparing documentation for grant applications, pre-registrations, or manuscript supplementary materials.

### Case 3: Preparing Multiple Scenarios

When preparing grant applications or pre-registrations, you might want to explore multiple scenarios (e.g., different effect size assumptions):

```{r workflow-scenarios}
# Define scenarios to compare
scenarios <- data.frame(
  effect = c(0.20, 0.25, 0.30),
  label = c("conservative", "expected", "optimistic")
)

# Generate reports for each scenario
for (i in seq_len(nrow(scenarios))) {
  plan_sample_size(
    f_expected = scenarios$effect[i],
    k_groups = 3,
    power = 0.90,
    output_dir = "sample_size_reports",
    output_file = sprintf("power_analysis_%s.html", scenarios$label[i]),
    open = FALSE,  # Don't open each one
    overwrite = TRUE
  )
}

message("Generated ", nrow(scenarios), " sample size reports")
```

This approach creates a set of reports that document your planning across different assumptions.


## Managing the Simulation Data

**Downloading Data Explicitly**

While `plan_sample_size()` downloads data automatically when needed, you can also download it explicitly:
```{r download-explicit}
# Download simulation data manually
download_sample_size_data()
```

This is useful if you want to:
- Pre-download data on a fast internet connection before traveling
- Verify the download completed successfully
- Troubleshoot download issues

To force a re-download (for example, after a package update with new simulation data):
```{r download-force}
download_sample_size_data(force = TRUE)
```

**Checking Cache Status**

To see whether data are cached and how much disk space they occupy:
```{r cache-status}
cache_info()
```

This displays:
- The cache directory location on your system
- Whether simulation data are currently cached
- The file size (approximately 15 MB when cached)

**Clearing the Cache**

If you need to free up disk space or suspect corrupted data, you can clear the cache:
```{r cache-clear}
cache_clear()
```

The data will be re-downloaded automatically the next time you run `plan_sample_size()`.

**Working with Simulation Data Directly**

Advanced users may want to access the raw simulation data for custom analyses or visualizations.
You can load the data directly into your R session:

```{r load-data-advanced}
# Load the complete simulation database
df_all <- load_sample_size_data()
```

This data frame contains all simulation results and can be filtered, summarized, or visualized using standard R tools.



