---
title: "Use Case"
author: "Meike Steinhilber"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 4
vignette: >
  %\VignetteIndexEntry{Use Case}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
csl: "apa.csl"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(sprtt)
df_stress <- df_stress[,3:4]
```

## Note

In the R code sections:

`# comment`: is a comment

`function()`: is R code

`#> results of function()`: is console output

## Overview

The `sprtt` package is the implementation of **s**equential **p**robability **r**atio **t**ests using the associated **t**-statistic (**sprtt**). This vignette describes an exemplary use case to improve the understanding of the package and the sequential t-test.

Other recommended vignettes cover:

-   [the theoretical background](https://meikesteinhilber.github.io/sprtt/articles/sequential_testing.html) and

-   a [general guide](https://meikesteinhilber.github.io/sprtt/articles/usage-sprtt.html), how to use the package.

## Use Case

A team of researchers wants to know if the stress of the employees increased in the company Test AG over the last year. The Human Resources department has gathered data from a new self-care tool they implemented for the employees. It is difficult to predict how many employees will still participate at the second measurement point (one year after the baseline). The team suggests using a sequential t-test instead of a Students t-test, because the sequential procedure can reduce the required sample size and is stopped right after the decision for one of the hypotheses. Furthermore, the sequential t-test is able to gather evidence for both the alternative and the null hypothesis.

### Hypothesis

The researchers know that the company has received more orders than in the year before. Thus, they do expect an increase in self-reported stress.

### Data Analysis

The parameters of the sequential t-test are specified as follows:

-   The researchers expect an increase in stress but they have no prior knowledge about the expected effect size. Thus, they define the lower limit of a substantial effect for their research interest.

    ```{r}
    d = 0.2
    ```

-   They choose the common alpha level of .05 and the same level for the beta error probability, which leads to a power of .95 (1 - ð›½).

    ```{r}
    alpha = 0.05
    power = 0.95
    ```

-   The data are repeated measures, therefore the test is a paired sequential t-test.

    ```{r}
    paired = TRUE
    ```

-   The researchers expect that the stress will have increased after one year. Thus, they specify that the true difference in means is greater than 0.\
    `(mean(one_year_stress) - mean(baseline_stress)) > 0`

    ```{r}
    alternative = "greater"
    ```

The HR department receives the new data piece by piece and passes it on directly to the researchers. The test is performed for the first time and starts with the first two data points.

```{r}
# first data from the Human Resources department -------------------------------
# current sample size
n_person = 2

# get data
df <- df_stress[1:n_person,]

# print data
df

# sequential t-test ------------------------------------------------------------
results <- seq_ttest(df$one_year_stress,
                     df$baseline_stress,
                     alpha = alpha,
                     power = power,
                     d = d,
                     paired = paired,
                     alternative = alternative)

# print results: entire output--------------------------------------------------
results
```

The decision from the first test is:

```{r}
results@decision
```

As a result, the researchers take one more data point and run the test again.

```{r}
# new data from the Human Resources department ---------------------------------
# get one more person
n_person = n_person + 1
df <- df_stress[1:n_person,]

# print new data
df

# sequential t-test ------------------------------------------------------------
results <- seq_ttest(df$one_year_stress,
                     df$baseline_stress,
                     alpha = alpha,
                     power = power,
                     d = d,
                     paired = paired,
                     alternative = alternative)

# print results ----------------------------------------------------------------
results@decision
```

This process is repeated until the decision is made for one of the two hypotheses. To simulate this sequential process, a while-loop embraces the sequential t-test function in the code below, which will not stop until the decision is to accept one of the hypotheses or the maximum of the data is reached.

```{r}
# define the starting point ----------------------------------------------------
decision = "continue sampling"
n_person = 3

# simulation of the sequential procedure ---------------------------------------
while(decision == "continue sampling") {
  # get the current data
  df <- df_stress[1:n_person,]
  
  # run the sequential test and save the results
  results <- seq_ttest(df$one_year_stress,
                       df$baseline_stress,
                       alpha = alpha,
                       power = power,
                       d = d,
                       paired = paired,
                       alternative = alternative)
  # save the current desicion
  decision = results@decision
  
  # add a new person
  n_person = n_person + 1
  
  # break if the maximum of the data is reached
  if (n_person > nrow(df_stress)) {
    break
  }
}

# console output ---------------------------------------------------------------
results
```

The while-loop comes to an end after the 48 data points.

### Report Results

```{r}
# Required results for the report ----------------------------------------------

# likelihood ratio (LR)
LR <- round(results@likelihood_ratio, digits = 2)
LR

# sample size (N)
N <- results@df + 2
N

# baseline stress (M and SD)
mean_t1 <- round(mean(df$baseline_stress), digits = 2)
mean_t1
sd_t1 <- round(sd(df$baseline_stress), digits = 2)
sd_t1

# after one year stress (M and SD)
mean_t2 <- round(mean(df$one_year_stress), digits = 2)
mean_t2
sd_t2 <- round(sd(df$one_year_stress), digits = 2)
sd_t2

# NOT IN THE PACKAGE INCLUDED --------------------------------------------------

# calculate effect size: CohenÂ´s d
d_results <- effsize::cohen.d(df$one_year_stress,
                 df$baseline_stress,
                 paired = TRUE)
d <- round(d_results$estimate, digits = 2)
d

# confidence intervall
d_lower <- round(d_results$conf.int[[1]], digits = 2)
d_lower
d_upper <- round(d_results$conf.int[[2]], digits = 2)
d_upper
```

Starting at *N* = 2, the test stops sampling at a total sample size of *N* = `r N` with *LR*~`r N`~ = `r LR`. This ratio indicates that the data are about 20 times more likely under H~1~ than under H~0~. Thus, we accept the alternative hypothesis: The perceived stress at the second measurement (*M* = `r mean_t2`, *SD* = `r sd_t2`) is higher than one year ago at the baseline measurement (*M* = `r mean_t1`, *SD* = `r sd_t1`), Cohen\`s *d* = `r d`, 95% CI [`r d_lower`, `r d_upper`].[^1]

[^1]: Note that this estimate of Cohen's d as well as the CI are based on the assumption of a fixed sample size and, thus, might be biased toward an overestimation of the true effect size. See @schnuerch2020a for details.

## References
